An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2021-07-09 latex_files/ch2/2_3_OMC.tex
added content in chapter 2

As already stated at Chapter~\ref{sec:introduction}, in
simulator-based models we cannot evaluate the posterior
$p(\thetab|\data) \propto L(\thetab)p(\thetab)$, due to the
intractability of the likelihood $L(\thetab) = p(\data|\thetab)$. The
following equation allows incorporating the simulator in the place of
the likelihood and forms the basis of all likelihood-free inference
approaches,

\begin{equation} \label{eq:likelihood}
  L(\thetab) =
  \lim_{\epsilon \to 0} c_\epsilon \int_{\yb \in B_{d,\epsilon}(\data)} p(\yb|\thetab)d\yb =
  \lim_{\epsilon \to 0} c_\epsilon \Pr(M_r(\thetab) \in B_{d,\epsilon}(\data))
\end{equation}
%
where $c_\epsilon$ is a proportionality factor dependent on
$\epsilon$, needed when
$\Pr(M_r(\thetab) \in B_{d, \epsilon}(\data)) \rightarrow 0$, as
$\epsilon \rightarrow 0$. Equation~\ref{eq:likelihood} describes that
the likelihood of a specific parameter configuration $\thetab$ is
proportional to the probability that the simulator will produce
outputs equal to the observations, using this configuration.

\subsubsection{Approximate Bayesian Computation (ABC) Rejection
  Sampling}

ABC rejection sampling is a modified version of the traditional
rejection sampling method, for cases when the evaluation of the
likelihood is intractable. In the typical rejection sampling, a sample
obtained from the prior $\thetab \sim p(\thetab)$ gets accepted
with probability $L(\thetab)/ \text{max}_{\thetab}
L(\thetab)$. Though we cannot use this approach out-of-the-box
(evaluating $L(\thetab)$ is impossible in our case), we can
modify the method incorporating the simulator.

In the discrete case scenario where $\Y_{\thetab}$ can take a finite
set of values, the likelihood becomes
$L(\thetab) = \Pr(\Y_{\thetab} = \data)$ and the posterior
$p(\thetab|\data) \propto \Pr(\Y_{\thetab}=\data)p(\thetab)$; hence, we can
sample from the prior $\thetab_i \sim p(\thetab)$, run the simulator
$\yb_i = M_r(\thetab_i)$ and accept $\thetab_i$ only if
$\yb_i = \data$.

The method above becomes less useful as the finite set of
$\Y_{\thetab}$ values grows larger, since the probability of
accepting a sample becomes smaller. In the limit
where the set becomes infinite (i.e.\ continuous case) the probability
becomes zero. In order for the method to work in this set-up, a
relaxation is introduced; we relax the acceptance criterion by letting
$\yb_{i}$ lie in a larger set of points i.e.\
$\yb_{i} \in \region(\data), \epsilon > 0$. The region can be
defined as $\region (\data) := \{\yb: d(\yb, \data) \leq \epsilon \}$
where $d(\cdot, \cdot)$ can represent any valid distance. With this
modification, the maintained samples follow the approximate posterior,

\begin{equation} \label{eq:approx_posterior}
  p_{d,\epsilon}(\thetab|\data) \propto Pr(\yb \in
  \region(\data)) p(\thetab)
\end{equation}

\noindent
This method is called Rejection ABC.

\subsubsection{Summary Statistics}

When $\yb \in \mathbb{R}^D$ lies in a high-dimensional space, generating
samples inside $\region (\data)$ becomes rare even when $\epsilon$ is
relatively large; this is the curse of dimensionality. As a
representative example lets make the following hypothesis;

\begin{itemize}
\item $d$ is set to be the Euclidean distance, hence
  $\region(\data) := \{ \yb: ||\yb - \data||_2^2 < \epsilon^2 \}$ is a
  hyper-sphere with radius $\epsilon$ and volume $V_{hypersphere} = \frac{\pi^{D/2}}{\Gamma(D/2 + 1)} \epsilon^D$
\item the prior $p(\thetab)$ is a uniform distribution in a hyper-cube with side of
  length $2\epsilon$ and volume $V_{hypercube} = (2\epsilon)^D$
\item the generative model is the identity function $\yb=f(\thetab)= \thetab $
\end{itemize}

\noindent
The probability of drawing a sample inside the hypersphere equals the
fraction of the volume of the hypersphere inscribed in the hypercube:

\begin{equation}
  Pr(\yb \in \region (\data))
  = Pr(\thetab \in \region (\data))
  = \frac{V_{hypersphere}}{V_{hypercube}}
  = \frac{\pi^{D/2}}{2^D\Gamma(D/2 + 1)} \rightarrow 0, \quad \text{as} \quad D \rightarrow \infty
\end{equation}

\noindent
We observe that the probability tends to $0$, independently of
$\epsilon$; enlarging $\epsilon$ will not increase the acceptance
rate. Intuitively, we can think that in high-dimensional spaces the
volume of the hypercube concentrates at its corners. This generates
the need for a mapping
$T: \mathbb{R}^{D_1} \rightarrow \mathbb{R}^{D_2}$ where $D_1 > D_2$,
for squeezing the dimensionality of the output. This
dimensionality-reduction step that redefines the area as
$\region(\data) := \{\yb: d(T(\yb), T(\data)) \leq \epsilon \}$ is
called \textit{summary statistic} extraction, since the distance is
not measured on the actual outputs, but on a summarisation (i.e.\
lower-dimension representation) of them.

\subsubsection{Approximations introduced so far}

So far, we have introduced some approximations for inferring the
posterior as
$p_{d,\epsilon}(\thetab|\data) \propto Pr(\Y_{\thetab} \in
\region(\data))p(\thetab)$ where
$\region(\data) := \{\yb: d(T(\yb), T(\data)) < \epsilon \}$. These
approximations introduce two different types of errors:

\begin{itemize}
\item $\epsilon$ is chosen to be \textit{big enough}, so that enough
  samples are accepted. This modification leads to the approximate
  posterior introduced in \eqref{eq:approx_posterior}
\item $T$ introduces some loss of information, making possible a $\yb$
  far away from $\data$ i.e.\ $\yb: d(\yb,\data)>\epsilon$, to enter
  the acceptance region after the dimensionality reduction
  $d(T(\yb), T(\data)) \leq \epsilon$
\end{itemize}

\noindent
In the following sections, we will not use the summary statistics in
our expressions for the notation not to clutter. One could understand
it as absorbing the mapping $T(\cdot)$ inside the simulator. In any
case, all the propositions that will be expressed in the following
sections are valid with the use of summary statistics.

\subsubsection{Optimisation Monte Carlo (OMC)}

@@

\begin{gather} \label{eq:approx_likelihood}
  L_{d, \epsilon}(\thetab) =
  \int_{\yb \in B_\epsilon(\data)}p(\yb|\thetab)d\yb =
  \int_{\yb \in \R^D} \indicator{\region(\data)}(\yb)p(\yb|\thetab)d\yb\\
  \approx \frac{1}{N} \sum_i^N \indicator{\region(\data)}(\yb_i),\text{ where }
  \yb_i \sim M_r(\thetab) \label{eq:init_view}\\
  \approx \frac{1}{N} \sum_i^N \indicator{\region (\data)} (\yb_i)
  \text{ where } \yb_i = M_d(\thetab, \vb_i), \vb_i \sim p(\vb) \label{eq:alt_view}
\end{gather}
%
This approach is quite intuitive; approximating the likelihood of a
specific $\thetab$ requires sampling from the data generator and count
the fraction of samples that lie inside the area around the
observations. Nevertheless, by using the approximation of equation
\eqref{eq:init_view} we need to draw $N$ new samples for each distinct
evaluation of $L_{d,\epsilon}(\thetab)$; this makes this approach
quite inconvenient from a computational point-of-view. For this
reason, we choose to approximate the integral as in equation
\eqref{eq:alt_view}; the nuisance variables are sampled once
$\vb_i \sim p(\vb)$ and we count the fraction of samples that lie
inside the area using the deterministic simulators
$M_d(\thetab, \vb_i) \: \forall i$. Hence, the evaluation
$L_{d,\epsilon}(\thetab)$ for each different $\thetab$ does not imply
drawing new samples all over again. Based on this approach, the
unnormalised approximate posterior can be defined as:

\begin{equation} \label{eq:aprox_posterior}
  p_{d,\epsilon}(\thetab|\data)
  \propto p(\thetab) \sum_i^N \indicator{ \region(\data)} (\yb_i)
\end{equation}

@@

The posterior approximation in \eqref{eq:approx_posterior} does not
provide any obvious way for drawing samples. In fact, the set
$\mathcal{S}_i = \{ \thetab: M_d(\thetab, \vb_i) \in \region(\data) \}$ can
represent any arbitrary shape in the D-dimensional Euclidean space; it
can be non-convex, can contain disjoint sets of $\thetab$ etc. We need
some further simplification of the posterior for being able to draw
samples from it.

As a side-note, weighted sampling could be performed in a
straightforward fashion with importance sampling. Using the prior as
the proposal distribution $\thetab_i \sim p(\thetab)$ and we can
compute the weight as
$w_i = \frac{L_{d,\epsilon}(\thetab_i)}{p(\thetab_i)}$, where
$L_{d,\epsilon}(\thetab_i)$ is computed with the expression
\eqref{eq:approx_likelihood}. This approach has the same drawbacks as
ABC rejection sampling; when the prior is wide or the dimensionality
$D$ is high, drawing a sample with non-zero weight is rare, leading to
either poor Effective Sample Size (ESS) or huge execution time.

>>>>

@@

\begin{gather} \label{eq:approx_likelihood_omc}
  L_{d, \epsilon}(\thetab) =
  \int_{\yb \in B_\epsilon(\data)}p(\yb|\thetab)d\yb =
  \int_{\yb \in \R^D} \indicator{\region(\data)}(\yb)p(\yb|\thetab)d\yb\\
  \approx \frac{1}{N} \sum_i^N \indicator{\region(\data)}(\yb_i),\text{ where }
  \yb_i \sim M_r(\thetab) \label{eq:init_view}\\
  \approx \frac{1}{N} \sum_i^N \indicator{\region (\data)} (\yb_i)
  \text{ where } \yb_i = M_d(\thetab, \vb_i), \vb_i \sim p(\vb) \label{eq:alt_view}
\end{gather}
%
This approach is quite intuitive; approximating the likelihood of a
specific $\thetab$ requires sampling from the data generator and count
the fraction of samples that lie inside the area around the
observations. Nevertheless, by using the approximation of equation
\eqref{eq:init_view} we need to draw $N$ new samples for each distinct
evaluation of $L_{d,\epsilon}(\thetab)$; this makes this approach
quite inconvenient from a computational point-of-view. For this
reason, we choose to approximate the integral as in equation
\eqref{eq:alt_view}; the nuisance variables are sampled once
$\vb_i \sim p(\vb)$ and we count the fraction of samples that lie
inside the area using the deterministic simulators
$M_d(\thetab, \vb_i) \: \forall i$. Hence, the evaluation
$L_{d,\epsilon}(\thetab)$ for each different $\thetab$ does not imply
drawing new samples all over again. Based on this approach, the
unnormalised approximate posterior can be defined as:

\begin{equation} \label{eq:approx_posterior_omc}
  p_{d,\epsilon}(\thetab|\data)
  \propto p(\thetab) \sum_i^N \indicator{ \region(\data)} (\yb_i)
\end{equation}

@@

The posterior approximation in \eqref{eq:approx_posterior_omc} does not
provide any obvious way for drawing samples. In fact, the set
$\mathcal{S}_i = \{ \thetab: M_d(\thetab, \vb_i) \in \region(\data) \}$ can
represent any arbitrary shape in the D-dimensional Euclidean space; it
can be non-convex, can contain disjoint sets of $\thetab$ etc. We need
some further simplification of the posterior for being able to draw
samples from it.

As a side-note, weighted sampling could be performed in a
straightforward fashion with importance sampling. Using the prior as
the proposal distribution $\thetab_i \sim p(\thetab)$ and we can
compute the weight as
$w_i = \frac{L_{d,\epsilon}(\thetab_i)}{p(\thetab_i)}$, where
$L_{d,\epsilon}(\thetab_i)$ is computed with the expression
\eqref{eq:approx_likelihood_omc}. This approach has the same drawbacks as
ABC rejection sampling; when the prior is wide or the dimensionality
$D$ is high, drawing a sample with non-zero weight is rare, leading to
either poor Effective Sample Size (ESS) or huge execution time.

### 2021-07-09 latex_files/ch2/2_4_ROMC.tex
added content in chapter 2

>>>>

\subsubsection*{\textit{Robust Optimisation Monte Carlo (ROMC) method}}

The ROMC method~\cite{Ikonomov2019} is very a recent likelihood-free
approach. Its fundamental idea is the transformation of the stochastic
data generation process $M_r(\thetab)$ to a deterministic function
$g_i(\thetab)$, by sampling the variables that produce the randomness
$\vb_i \sim p(\V)$. Formally, in every stochastic process the
randomness is influenced by a vector of random variables $\V$, whose
state is unknown prior to the execution of the simulation; sampling in
advance this state makes the procedure deterministic, i.e.
$g_i(\thetab) = M_d(\thetab, \V=\vb_i)$. This approach was initially
introduced by \cite{Meeds2015} as \textit{Optimisation Monte Carlo
  (OMC)}. The ROMC extended this approach by resolving a fundamental
failure-mode of OMC. The ROMC describes a methodology for
approximating the posterior through a series of steps, without
explicitly enforcing which algorithms must be utilised for each
step\footnote{The implementation chooses a specific algorithm for each
  task, but this choice has just a demonstrative value; any
  appropriate algorithm can be used instead.}; in this sense, it can
be perceived as a meta-algorithm.

### 2021-07-09 latex_files/ch3/3_2_1_general_design.tex
added content in chapter 2

For illustrating the functionalities we choose as running example the
following model, introduced by~\autocite{Ikonomov2019},

@@

\begin{figure}[h]
    \begin{center}
\includegraphics[width=0.7\textwidth]{./Thesis/images/chapter3/example_gt.png}
    \end{center}
  \caption[Ground-truth posterior distribution of the simple 1D
example.]{Ground-truth posterior distribution of the simple 1D
example.}
  \label{fig:example_gt}
\end{figure}

@@

\begin{figure}[!ht]
    \begin{center}
\includegraphics[width=0.8\textwidth]{./Thesis/graphs/ROMC.png}
    \end{center}
    \caption[Overview of the ROMC implementation.]{Overview of the
ROMC implementation. The training part follows a sequential pattern;
the functions in the green ellipses must be called in a sequential
fashion for completing the training part and define the posterior
distribution. The functions in blue ellipses are the functionalities
provided to the user.}
    \label{fig:romc_overview}
\end{figure}

>>>>

For illustrating the functionalities we choose as running example the
following model, introduced by~\cite{Ikonomov2019},

@@

\begin{figure}[h]
    \begin{center}
\includegraphics[width=0.7\textwidth]{./latex_files/images/chapter3/example_gt.png}
    \end{center}
  \caption[Ground-truth posterior distribution of the simple 1D
example.]{Ground-truth posterior distribution of the simple 1D
example.}
  \label{fig:example_gt}
\end{figure}

@@

\begin{figure}[!ht]
    \begin{center}
\includegraphics[width=0.8\textwidth]{./latex_files/graphs/ROMC.png}
    \end{center}
    \caption[Overview of the ROMC implementation.]{Overview of the
ROMC implementation. The training part follows a sequential pattern;
the functions in the green ellipses must be called in a sequential
fashion for completing the training part and define the posterior
distribution. The functions in blue ellipses are the functionalities
provided to the user.}
    \label{fig:romc_overview}
\end{figure}

### 2021-07-09 latex_files/ch3/3_2_2_training.tex
added content in chapter 2

\begin{enumerate}[label=(\roman*)]
\item \mintinline{python}{romc.solve_problems(n1, use_bo=False, optimizer_args=None, seed=None)}
\item \mintinline{python}{romc.estimate_regions(eps_filter,}

@@

\begin{figure}[h]
    \begin{center}
      \includegraphics[width=0.48\textwidth]{./Thesis/images/chapter3/example_theta_dist.png}
      \includegraphics[width=0.48\textwidth]{./Thesis/images/chapter3/example_region.png}
    \end{center}
  \caption[Histogram of distances at the 1D example.]{Histogram of distances and visualisation of a specific region.}
  \label{fig:example_training_hist}
\end{figure}

>>>>

\begin{enumerate}
\item \mintinline{python}{romc.solve_problems(n1, use_bo=False, optimizer_args=None, seed=None)}
\item \mintinline{python}{romc.estimate_regions(eps_filter,}

@@

\begin{figure}[h]
    \begin{center}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter3/example_theta_dist.png}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter3/example_region.png}
    \end{center}
  \caption[Histogram of distances at the 1D example.]{Histogram of distances and visualisation of a specific region.}
  \label{fig:example_training_hist}
\end{figure}

### 2021-07-09 latex_files/ch3/3_2_3_inference.tex
added content in chapter 2

\begin{enumerate}[label=(\roman*)]
\item \mintinline{python}{romc.sample(n2, seed=None)}
\item \mintinline{python}{romc.compute_expectation(h)}  
\item \mintinline{python}{romc.eval_unnorm_posterior(theta)}
\item \mintinline{python}{romc.eval_posterior(theta)}
\end{enumerate}

@@

\begin{figure}[h]
    \begin{center}
      \includegraphics[width=0.48\textwidth]{./Thesis/images/chapter3/example_marginal.png}
      \includegraphics[width=0.48\textwidth]{./Thesis/images/chapter3/example_region_samples.png}
    \end{center}
  \caption[Histogram of the obtained samples at the 1D example.]{(a) Left: Histogram of the obtained samples. (b) Right: Acceptance region around $\theta_1^*$ with the obtained samples plotted inside.}
  \label{fig:example_sampling}
\end{figure}

@@

\begin{figure}[ht]
    \begin{center}
      \includegraphics[width=0.75\textwidth]{./Thesis/images/chapter3/example_posterior.png}
    \end{center}
  \caption[Approximate posterior evaluation, at the 1D example.]{Approximate posterior evaluation.}
  \label{fig:approx_posterior}
\end{figure}

>>>>

\begin{enumerate}
\item \mintinline{python}{romc.sample(n2, seed=None)}
\item \mintinline{python}{romc.compute_expectation(h)}  
\item \mintinline{python}{romc.eval_unnorm_posterior(theta)}
\item \mintinline{python}{romc.eval_posterior(theta)}
\end{enumerate}

@@

\begin{figure}[h]
    \begin{center}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter3/example_marginal.png}
      \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter3/example_region_samples.png}
    \end{center}
  \caption[Histogram of the obtained samples at the 1D example.]{(a) Left: Histogram of the obtained samples. (b) Right: Acceptance region around $\theta_1^*$ with the obtained samples plotted inside.}
  \label{fig:example_sampling}
\end{figure}

@@

\begin{figure}[ht]
    \begin{center}
      \includegraphics[width=0.75\textwidth]{./latex_files/images/chapter3/example_posterior.png}
    \end{center}
  \caption[Approximate posterior evaluation, at the 1D example.]{Approximate posterior evaluation.}
  \label{fig:approx_posterior}
\end{figure}
